Supporting Information: Improving cropland maps through tight integration of human and machine intelligence

### 2020-08-05 inst/manuscript/figures.Rmd
Completed figure and text for segmentation accuracy assessment. Some reordering of text in methods and results.

leglabs <- seq(0.4, 1, 0.1)
p1 <- ggplot(africa) + geom_sf(fill = "gray97", color = "grey", size = 0.25) +
  geom_sf(data = tiles, fill = "grey85", col = "grey", size = 0.3) +
  geom_sf(data = ghana, fill = "transparent", col = "grey", size = 0.3) +
  geom_sf(data = image_quality$tile, 
          mapping = aes(fill = Score, color = Score)) +
  facet_wrap(vars(season), ncol = 2) +
  scale_fill_gradientn(colors = pal(7), 
                       breaks = seq(0.4, 1, 0.1), limits = c(0.4, 1),
                       labels = leglabs) + 
  scale_color_gradientn(colors = pal(7), 
                        breaks = seq(0.4, 1, 0.1), limits = c(0.4, 1), 
                        labels = leglabs) + 
  coord_sf(xlim = maprange[1:2], ylim = maprange[3:4]) + 
  theme_linedraw() + 
  theme(legend.position="bottom", 
        strip.background = element_rect(fill = "white"), 
        strip.text = element_text(colour = "black"), 
        panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

@@

>>>>

leglabs <- seq(0.4, 1, 0.1)
p1 <- ggplot(africa) + geom_sf(fill = "gray97", color = "grey", size = 0.25) +
  geom_sf(data = tiles, fill = "grey85", col = "grey", size = 0.3) +
  geom_sf(data = ghana, fill = "transparent", col = "grey", size = 0.3) +
  geom_sf(data = image_quality$tile, 
          mapping = aes(fill = Score, color = Score)) +
  facet_wrap(vars(season), ncol = 2) +
  scale_fill_gradientn(colors = pal(7), 
                       breaks = seq(0.4, 1, 0.1), limits = c(0.4, 1),
                       labels = leglabs) + 
  scale_color_gradientn(colors = pal(7), 
                        breaks = seq(0.4, 1, 0.1), limits = c(0.4, 1), 
                        labels = leglabs) + 
  coord_sf(xlim = maprange[1:2], ylim = maprange[3:4]) + 
  theme_linedraw() + 
  theme(legend.position = "bottom", 
        strip.background = element_rect(fill = "white"), 
        strip.text = element_text(colour = "black"), 
        panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text = element_blank())

@@

### Figure 6: Categorical map accuracy

@@

### Figure S7: Segmentation statistics
```{r segstats, eval=FALSE}
data("top_label_data")
data("field_validation_stats")

human <- top_label_data$stats %>%
  mutate(mu_prop = NA, type = "Validation") %>%
  select(-contains("id"))
machine <- field_validation_stats %>%
  mutate(type = "Segmentation")
humachine <- bind_rows(human, machine) %>% 
  mutate(aoi = as.numeric(gsub("labeller", "", aoi))) %>% 
  arrange(aoi)

# Summarize AOI-wise averages (field size)
humachine_sum <- humachine %>% group_by(aoi, type) %>% 
  summarize(Median = median(Mean), Mean = mean(Mean)) %>% 
  ungroup() %>% 
  tidyr::pivot_longer(cols = Median:Mean, names_to = "Stat")
# field number
humachine_sum_n <- humachine %>% group_by(aoi, type) %>% 
  summarize(Median = median(n), Mean = mean(n)) %>% 
  ungroup() %>% 
  tidyr::pivot_longer(cols = Median:Mean, names_to = "Stat")

# plots
fcols <- c(Segmentation = rgb(0, 0, 1, alpha = 0.5),
          Validation = rgb(1, 0, 0, alpha = 0.5))
brks <- seq(0, 30, 5)
theme_fun <- function(x) {
  theme_linedraw() + 
  theme(legend.position = "bottom", legend.text = element_text(size = 11),
        strip.background = element_rect(fill = "white"), 
        strip.text = element_text(size = 11, colour = "black"),  
        axis.text.x = element_text(size = 11, hjust = 1, vjust = 0.5, 
                                   angle = 90), 
        axis.text.y = element_text(size = 11))
}

# Field size
p <- ggplot(humachine) + 
  geom_density(aes(x = Mean, y = ..count.., fill = type), col = "transparent", 
               linetype = "blank") +  
  geom_vline(data = humachine_sum, 
             aes(xintercept = value, color = type, linetype = Stat), 
             size = 0.5) +
  scale_x_continuous(expand = c(0.03, 0.03), breaks = brks, limits = c(0, 30)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_color_manual(name = "", values = c("blue", "red"), guide = FALSE) + 
  scale_fill_manual(name = "", values = fcols) +  
  # scale_linetype_manual(name = "", values = c("dashed", "solid")) +  
  scale_linetype("") +  
  facet_wrap(vars(aoi)) + 
  xlab("Field size (ha)") + ylab("Count") + 
  theme_fun()
ggsave(pth("figures/si_validation_stats_fha.png"), 
       width = 7, height = 7, units = "in", dpi = 300)

# field number
brks <- seq(0, 20, 5)
p2 <- ggplot(humachine) + 
  geom_density(aes(x = n, y = ..count.., fill = type), col = "transparent", 
               linetype = "blank") +  
  geom_vline(data = humachine_sum_n, 
             aes(xintercept = value, color = type, linetype = Stat), 
             size = 0.5) +
  scale_x_continuous(expand = c(0.03, 0.03), breaks = brks, limits = c(0, 20)) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 40)) +
  scale_color_manual(name = "", values = c("blue", "red"), guide = FALSE) + 
  scale_fill_manual(name = "", values = fcols) +  
  # scale_linetype_manual(name = "", values = c("dashed", "solid")) +  
  scale_linetype("") +  
  facet_wrap(vars(aoi)) + 
  xlab("Number of fields") + ylab("Count") + 
  theme_fun()
ggsave(pth("figures/si_validation_stats_fn.png"), 
       width = 7, height = 7, units = "in", dpi = 300)

### 2020-08-05 inst/manuscript/manuscript.Rmd
Completed figure and text for segmentation accuracy assessment. Some reordering of text in methods and results.

library(knitr)
library(citr)
library(dplyr)

@@

#### Feature extraction
The first step is the extraction of additional features from the growing season/non-growing season PlanetScope imagery. Previous work showed that a large number of simple features that summarize the statistical properties of reflectance and vegetatiation indices in local neighborhoods are highly effective for classifying smallholder croplands [@Debatsgeneralizedcomputervision2016]. We followed this logic in this study, but opted for a smaller feature space to accomodate the storage and memory requirements of mapping geography that were several orders of magnitude larger. Here, we used a set of 24 features, comprising the eight original composite bands (four from each season), an 11X11 mean of each band, as well as the standard deviation of each band within a 5X5 window (Table 1).

@@

After fitting, the model is applied to the features of the model validation set, and a set of accuracy metrics are calculated, including ordinary binary accuracy, the F1 score, and the area under the curve of the Receiver Operating Characteristic [@Pontiustotaloperatingcharacteristic2014].

@@

### Image quality assessment
```{r, echo=FALSE}
data("image_quality")
tile_quality <- image_quality %>% filter(kml_type == "F") %>% 
  group_by(name, season, worker_id) %>% 
  summarize(tile = min(tile), Score = sum(score) / 12) %>% ungroup() 
tilen <- tile_quality %>% distinct(tile) %>% count() %>% pull(n)
```

@@

### Accuracy assessment

The accuracy assessment procedure described in section 2.2 was used strictly to assess performance gains in the active learning loop, according to suggested  best practices [@ElmesAccountingtrainingdata2020]. To independently assess the categorical accuracy of final map products, we followed guidelines provided by [@StehmanKeyissuesrigorous2019] to create a map reference sample. To develop this sample, we first made a stratified random sample of grid cells across the entire country, using the mapped field boundaries to define the two strata. Next, a primary supervisor examined the image composites in each randomly selected cell, and placed a square polygon of ~0.1 ha on the primary class intersecting the centerpoint of the cell. The square was shifted so that it was contained entirety with the intersected class: cropland or non-cropland. Where the supervisor was uncertain about the class label, one of two additional classes were selected: likely cropland or likely non-cropland. After placing these polygons, the supervisor removed their classes from the polygons and sent them to three separate experts, who labelled the polygons according to their interpretation of the underlying imagery. The four sets of labels were then combined and the overall agreement between them was calculated. This agreement provided the upper bound on knowable map accuracy. See the SI for further details on the creation and analysis of the map reference sample.

To assess the accuracy of the segmented field boundaries, we compared the size class distributions of the segmented field boundaries against those of the  workers' digitized polygons at map validation sites. We calculated the differences in the measures of central tendency (mean, median, mode) between the two datasets within each validation site, and at the scale of the AOI.

To further evaluate the capacities of different components of the system, we performed several further experiments. In the first experiment, we evaluated the degree to which training label quality impacts final map accuracy, by comparing maps created by training the RandomForest algorithm with labels created by the least accurate worker, the most accurate worker, as well as those trained with the consensus labels. This test aligns with Tier 1 standards for accounting for the impact of training data error on final map accuracy [@ElmesAccountingtrainingdata2020]. Next, we evaluated the degree to which training sample selection via active learning impacted map accuracy relative to random selection of additional training sites. Lastly, we repeated the entire map development and map assessment process for one AOI for the 2019 agricultural year, in order to assess the ability of the system to create annually updated maps and detect inter-annual change.

@@

# Results
```{r, echo=FALSE, message=FALSE, warning=FALSE}
tiles <- sf::read_sf(system.file("extdata/ghana_tiles.geojson", 
                                 package = "activemapper"))
km2 <- round(as.numeric(units::set_units(sum(sf::st_area(tiles)), "km2")))
km2 <- format(km2, big.mark = ",", scientific = FALSE)
```
We developed maps for all of Ghana (`r km2` km$^2$). Ghana has several agro-ecologically distinct regions, ranging from the primarily grain and vegetable crop producing regions in the northern savannas to tree crop-dominated system in the forested southwest, where cocoa and oil palm are among the dominant crops. For these latter regions, we did not attempt to classify tree crops, and instead mapped clearings that potentially contain field crops or recently felled or newly replanted tree crops. Our rationale for this decision was that the resolution of PlanetScope imagery is insufficient to distinguish such crops from the surrounding forest, and clear field boundaries are typically not visible in cocoa crops.

## Image catalog
```{r, echo=FALSE}
tqual_mu <- image_quality$tile %>%
  group_by(season, tile) %>% 
  summarize(score = mean(Score)) %>% 
  ungroup()
tqual_sum <- tqual_mu %>% 
  group_by(season) %>% 
  summarize(mu = mean(score), sd = sd(score))
tqmu <- tqual_sum %>% pull(mu) %>% round(2)
names(tqmu) <- tqual_sum %>% pull(season)

@@

### Model performance
```{r, echo=FALSE}
data("iteration_metrics")
delta_func <- function(x, rnd = 4) round((x - lag(x)) / lag(x), rnd)

@@

@@

### Segmentation accuracy
To evaluate the accuracy of the segmented field boundaries, we calculate the mean and standard deviation of the most accurate (for the given AOI) labeller's maps for validation sites, calculated the mean and standard deviation of area, and compared these statistics to those derived from the model-generated segments for those sites (Figure X).

>>>>

library(knitr)
library(citr)
library(dplyr)
library(activemapper)

@@

#### Feature extraction
The first step is the extraction of additional features from the growing season/non-growing season PlanetScope imagery. Previous work showed that a large number of simple features that summarize the statistical properties of reflectance and vegetatation indices in local neighborhoods are highly effective for classifying smallholder croplands [@Debatsgeneralizedcomputervision2016]. We followed this logic in this study, but opted for a smaller feature space to accomodate the storage and memory requirements of mapping geography that were several orders of magnitude larger. Here, we used a set of 24 features, comprising the eight original composite bands (four from each season), an 11X11 mean of each band, as well as the standard deviation of each band within a 5X5 window (Table 1).

@@

After fitting, the model is applied to the features of the model validation set, and a set of performance metrics are calculated, including ordinary binary accuracy, the F1 score (the geometric mean of user's and producer's accuracy), and the area under the curve of the Receiver Operating Characteristic [@Pontiustotaloperatingcharacteristic2014].

@@

### Image quality
```{r, echo=FALSE}
data("image_quality")
tile_quality <- image_quality %>% filter(kml_type == "F") %>% 
  group_by(name, season, worker_id) %>% 
  summarize(tile = min(tile), Score = sum(score) / 12) %>% ungroup() 
tilen <- tile_quality %>% distinct(tile) %>% count() %>% pull(n)
```

@@

### Map assessment
#### Validation performance
To assess the gain in RandomForest performance due to active learning loop, we assessed the change in accuracy, F1, and AUC (see 2.2.3.2) between each iteration and between the first and last iterations for each AOI. We then compared the magnitude of these iteration-wise changes to those that occurred when each batch of new training sites was randomly selected [@DebatsIntegratingactivelearning2017].

#### Independent accuracy assessment
The accuracy assessment procedure described above was not strictly independent because it was re-used the same validation sites over multiple iterations [@ElmesAccountingtrainingdata2020]. To independently assess the accuracy of final map products, we followed guidelines provided by [@StehmanKeyissuesrigorous2019] to create a separate map reference sample. To develop this sample, we first used the segmented field boundaries to define a cropland and non-cropland stratum for Ghana, and then randomly selected a sample of the centroids of grid cells intersecting each stratum. Next, a primary supervisor examined the image composites in each randomly selected cell, and placed a square polygon of ~0.1 ha on the primary class intersecting the centerpoint of the cell. The square was shifted so that it was contained entirety with the intersected class: cropland or non-cropland. Where the supervisor was uncertain about the class label, one of two additional classes were selected: likely cropland or likely non-cropland. After placing these polygons, the supervisor removed their classes from the polygons and sent them to three separate experts, who labelled the polygons according to their interpretation of the underlying imagery. The four sets of labels were then combined and the overall agreement between them was calculated. This agreement provided the upper bound on knowable map accuracy. See the SI for further details on the creation and analysis of the map reference sample. These points were then intersected with both the RandomForest probability images and the segmented field boundaries, and confusion matrixes between the map reference sample labels and the extracted map classes was constructed to assess the categorical accuracy of each map product.

To assess the accuracy of the segmented field boundaries, we compared the size class distributions of the segmented field boundaries against those of the  workers' digitized polygons at map validation sites. We selected the field polygons from the most accurate labeller to digitize each of the 100 validation sites in each AOI, and calculated the site-wise average area and number of polygons. We then calculated the same statistics from the segmented boundaries that intersected each validation grid. We compared the distributions and proximity of two measures of central tendency (mean and median) calculated from the two datasets for each AOI, and across all AOIs.

## The impact of training data error
To quantify the potential deleterious impact of training data error on our map results, we compared maps created by training the RandomForest algorithm with three sets of labels: 1) those from the lowest scoring worker to map each training site, 2) those from the most accurate worker, and 3) those trained as usual with the consensus labels. We then quantified the prediction performance against the validation sites, and compared the differences in performance metrics for each variant. This assessment follows recommended Tier 1 (i.e. best practice) standards to account for training data errors in machine learning-based Earth Observation applications [@ElmesAccountingtrainingdata2020].

```{r, echo=FALSE}
load(system.file("extdata/", "instance_tbls.rda", package = "activemapper"))
kml_types <- instance_tbls$kml_data %>% filter(aoi == "labeller8v2") %>% 
  group_by(kml_type) %>% count() %>% filter(kml_type == "Q") %>% pull(n)
```

As a final test of training data quality on map accuracy, we retrained the model for AOI 8 for 2018 using new labels that were generated against an updated training reference library (n = `r kml_types`), and compared the performance of this model against the previous version of the model.

@@

# Results
```{r, echo=FALSE, message=FALSE, warning=FALSE}
tiles <- sf::read_sf(system.file("extdata/ghana_tiles.geojson", 
                                 package = "activemapper"))
km2 <- round(as.numeric(units::set_units(sum(sf::st_area(tiles)), "km2")))
km2 <- format(km2, big.mark = ",", scientific = FALSE)
```
We developed maps for all of Ghana (`r km2` km$^2$). Ghana has several agro-ecologically distinct regions, ranging from the primarily grain and vegetable crop producing regions in the northern savannas to tree crop-dominated system in the forested southwest, where cocoa and oil palm are among the dominant crops. For these latter regions, we did not attempt to classify tree crops, and instead mapped clearings that potentially contain field crops or newly felled or replanted tree crops. We made this decision because the resolution of PlanetScope imagery is insufficient to distinguish tree crops from the surrounding forest, and clear field boundaries are typically not visible in cocoa crops.

## Image catalog and quality
```{r, echo=FALSE}
data("image_quality")
tqual_mu <- image_quality$tile %>%
  group_by(season, tile) %>% 
  summarize(score = mean(Score)) %>% 
  ungroup()
tqual_sum <- tqual_mu %>% 
  group_by(season) %>% 
  summarize(mu = mean(score), sd = sd(score))
tqmu <- tqual_sum %>% pull(mu) %>% round(2)
names(tqmu) <- tqual_sum %>% pull(season)

@@

### Validation performance
```{r, echo=FALSE}
data("iteration_metrics")
delta_func <- function(x, rnd = 4) round((x - lag(x)) / lag(x), rnd)

@@

We compared the gain in performance at AOIs 1, 8, and 15 against models trained for those same locations using the initial random draw of 500, followed by increments of 100 randomly selected points per iteration.

@@

### Segmentation accuracy
```{r, echo=FALSE}
# validation statistics
data("top_label_data")
data("field_validation_stats")

human <- top_label_data$stats %>%
  mutate(mu_prop = NA, type = "val") %>%
  select(-contains("id"))
machine <- field_validation_stats %>%
  mutate(type = "seg")
humachine <- bind_rows(human, machine) %>% 
  mutate(aoi = as.numeric(gsub("labeller", "", aoi))) %>% 
  arrange(aoi)

huma_aois <- humachine %>% group_by(aoi, type) %>% 
  summarize(med = median(Mean), mu = mean(Mean)) %>% 
  ungroup()
lrgst <- huma_aois %>% select(-med) %>% filter(aoi %in% c(2, 3, 7, 15)) %>% 
  tidyr::pivot_wider(names_from = type, values_from = "mu") %>% 
  filter(seg %in% range(seg))
lrgstlow <- lrgst %>% filter(seg == min(seg))
lrgsthigh <- lrgst %>% filter(seg == max(seg))

# overall stats
huma_sum <- humachine %>% group_by(type) %>% 
  summarize(med = median(Mean), sd = sd(Mean), min = min(Mean), 
            max = max(Mean), mu = mean(Mean)) %>% 
  select(type, mu, med, sd, min, max)
segha <- huma_sum %>% filter(type == "seg") %>% pull(mu) %>% round(2)
valha <- huma_sum %>% filter(type == "val") %>% pull(mu) %>% round(2)
seghamed <- huma_sum %>% filter(type == "seg") %>% pull(med) %>% round(2)
valhamed <- huma_sum %>% filter(type == "val") %>% pull(med) %>% round(2)
seghasd <- huma_sum %>% filter(type == "seg") %>% pull(sd) %>% round(2)
valhasd <- huma_sum %>% filter(type == "val") %>% pull(sd) %>% round(2)

huma_sumn <- humachine %>% group_by(type) %>% 
  summarize(med = median(n), sd = sd(n), min = min(n), max = max(n), 
            mu = mean(n)) %>% 
  select(type, mu, med, sd, min, max)
segn <- huma_sumn %>% filter(type == "seg") %>% pull(mu) %>% round(2)
valn <- huma_sumn %>% filter(type == "val") %>% pull(mu) %>% round(2)
segnmed <- huma_sumn %>% filter(type == "seg") %>% pull(med) %>% round(2)
valnmed <- huma_sumn %>% filter(type == "val") %>% pull(med) %>% round(2)
segnsd <- huma_sumn %>% filter(type == "seg") %>% pull(sd) %>% round(2)
valnsd <- huma_sumn %>% filter(type == "val") %>% pull(sd) %>% round(2)
```
The comparisons of digitized versus segmented field boundaries showed that the mean field size across all 1600 validation sites averaged `r round(segha, 2)` ha (StDev = `r round(seghasd, 2)`), which was `r round((segha - valha) / valha * 100) / 100` times larger than the `r round(valha, 2)` ha (StDev = `r round(valhasd, 2)`) mean area of labeller-digitized polygons. This discrepancy was primarily caused by segmentation results in four AOIs (2, 3, 7, and 15; Figure S5), where segments averaged between `r round(lrgstlow$seg, 2)` and `r round(lrgsthigh$seg, 2)` ha, compared to `r round(lrgstlow$val, 2)` - `r round(lrgsthigh$val, 2)` for the corresponding hand-digitized polygons. In contrast, the median segmented field area was closer to that of digitized polygons (`r seghamed` for segments versus `r valhamed` for labeller polygons). The number of segmented fields per validation site averaged `r segn` (median = `r segnmed`; StDev = `r segnsd`) compared to `r valn` (median = `r valnmed`; StDev = `r valnsd`) for digitized polygons (Figure S6).

### 2020-08-05 inst/manuscript/si.Rmd
Completed figure and text for segmentation accuracy assessment. Some reordering of text in methods and results.

The distribution of data collection effort and scores received at training reference sites by labelling team (Figure \ref{fig:trainqual}). These results include those from a repeated labelling of the initial training in Cluster 2.  This was done because we discovered a small spatial offset in the original image composites, which we corrected by reprocessing the images. We replaced the image overlays in the instance for Cluster 2 \ref{fig:trainval}A), and the labelling team mapped these sites a second time during the production run in late 2019. The reprocessed labels were used to initially train the model for AOIs 9, 12, and 15.

@@

>>>>

The distribution of data collection effort and scores from training reference sites by labelling team (Figure \ref{fig:trainqual}). These results include those from repeat labelling of the initial training in Cluster 2. This was done because we discovered a small spatial offset in the original image composites, which we corrected by reprocessing the images. We replaced the image overlays in the instance for Cluster 2 \ref{fig:trainval}A), and the labelling team mapped these sites a second time during the production run in late 2019. The reprocessed labels were used to initially train the model for AOIs 9, 12, and 15.

@@

## Map accuracy 
### Categorical
[To fill]

### Field area and number
To mean, median, and distributions of the average area of segmented field boundaries over the 100 validation sites in each AOI are compared to the areas of the polygons digitized by the most accurate labeller over the same sites in Figure \ref{fig:areavalidation}. The same statistics for average number of segments versus average number of labelled polygons across validation sites in each AOI are shown in Figure \ref{fig:numbervalidation}.      
```{r areavalidation, echo = FALSE, out.width="100%", fig.cap="The distributions of the average areas (in hectares) of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average areas of field boundaries digitized (shown in red) by the most accurate worker to label each site. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_validation_stats_fha.png')
```

```{r numbervalidation, echo = FALSE, out.width="100%", fig.cap="The distributions of the average number of segmented field boundaries (shown in blue) at the 100 validation sites per AOI, compared with the average number of digitized polygons (shown in red) by the most accurate worker to label each site. The mean and median of each distribution is shown. Vertical lines indicate the mean and median of each distribution.", fig.align='center', message=FALSE, results='hold'}
knitr::include_graphics('figures/si_validation_stats_fn.png')
```
